---
title: "Introduction to R"
subtitle: "Practice exercises"
author: "Kim Dill-McFarland, <kadm@uw.edu>"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: console
---
```{r include=FALSE}
knitr::opts_chunk$set(fig.width = 4, fig.height = 3)
```

# Setup

Open the Intro R Rproject and start a new working script. Install any new packages, load packages, set a seed, and load data as we did in the workshop.

```{r message=FALSE}
#RNAseq expression and metadata
load("data/RSTR_data_clean_subset.RData")
```

# Exercises
## Day 1: Base R
### Project setup

1. What are the benefits of storing data in `RData` versus tables (`csv`, `tsv`, etc)?
1. Imagine a hypothetical project with the following data and results. How would you choose to setup your Rproject directory and sub-directories? This is something that may evolve over time, but it is helpful to start with a defined structure to make it easier for you and others to find things.
    - `.RData` file containing all cleaned data for the project
    - 2 `.csv` of raw RNAseq counts and sample metadata (what was cleaned to make the `.RData`)
    - 4 `.csv` with linear model results
    - 25 `.png` plots of gene expression, individual genes
    - 1 `.png` plot of gene expression, faceted with many genes
    - 2 `.R` scripts, 1 for linear modeling and 1 for making plots
    - 1 `.Rmd` report summarizing and interpreting the results
    
### Data types

1. What is the difference between a `character` and `facter`?
1. What data type does R classify the date `2021.06`? What about `2021/06`? If it is not classified as a "date", how could this impact downstream analyses? Try to predict the outcomes before checking in R. 
    - Challenge: Checkout the package `lubridate` for functions to effectively work with dates in R.
1. You have an `S3` list object named `myData`and it contains 2 data frames named `A` and `B`. Within `B` there is a column named `variable1`. How do you access this variable?

### Subsetting and filtering

Using `dat`:

1. What is the mean library size `lib.size`? 
1. Try running `summary(dat$targets)`. What kinds of data does it provide? Why are the results different for different variables?
1. How many libraries have a library size `lib.size` greater than 5 million and a normalization factor `norm.factors` less than 1?
1. Challenge: Using the function `grepl`, how many libraries are from a donor with an `RSID` that starts with "RS1025"?

# R session

```{r}
sessionInfo()
```

***