---
title: "Data visualization in ggplot"
subtitle: "RSTR-seq edition"
author: "Kim Dill-McFarland"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: console
urlcolor: blue
fig_width: 6 
---

```{r, include=FALSE}
#set working dir to project
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# 3. Data visualization in ggplot

In this workshop, we introduce you to data visualization in `ggplot` ([docs.ggplot2.org](http://docs.ggplot2.org/current/)), the [tidyverse](https://www.tidyverse.org/) package for plots. In it, we cover how to create:

* Volcano plot
* PCA plot
* Customization such as color, shape, theme
* Facets

During the workshop, we will build an R script together, which will be posted as 'live_notes' after the workshop at <https://github.com/BIGslu/workshops/tree/main/2022.08.15_R.tidyverse.workshop/live_notes>

We load `ggplot` within the `tidyverse.` For more information on installation, see the [setup instructions][lesson0].

```{r}
library(tidyverse)
```

## Intro to ggplot

### Why ggplot?

ggplot2 is an implementation of _Grammar of Graphics_ (Wilkinson 1999) for R

Benefits:

- handsome default settings
- snap-together building block approach
- automatic legends, colors, facets
- statistical overlays like regressions lines and smoothers (with confidence intervals)

Drawbacks:

- it can be hard to get it to look *exactly* the way you want
- requires having the input data in a certain format

### ggplot layers

`ggplot` functions by adding layers to a plot, somewhat like how `dplyr`/`tidyr` sequentially modify data with the pipe `%>%`. Instead, `ggplot` uses the `+` to combine layers such as:

- _data_: 2D table (`data.frame`) of _variables_
- _aesthetics_: map variables to visual attributes (position, color, etc)
- _geoms_: graphical representation of data (points, lines, etc)
- _stats_: statistical transformations to get from data to points in
the plot (binning, summarizing, smoothing)
- _scales_: control how to map a variable to an aesthetic
- _facets_: juxtapose mini-plots of data subsets, split by variables
- _guides_: axes, legend, etc. that reflect the variables and their values

The idea is to independently specify and combine layers to create the plot you want. There are at least three things we have to specify to create any plot:

1. data
2. aesthetic mappings from data variables to visual properties
3. a geom describing how to draw those properties

## Data

A quick reminder about our data. There are bulk RNA-seq contained in the `dat` object in file "data/dat_voom.RData". Normalized log2 counts per million (CPM) are contained in `dat$E` and library metadata are in `dat$targets`. You will not need the other pieces of `dat` for this workshop.

```{r}
load("data/dat_voom.RData")

dat$E[1:3,1:3]

dat$targets[1:3,]
```

Here is a brief description of the variables in the metadata.

* `libID` (character): Unique library ID for each sequencing library. Formatted as ptID_condition
* `ptID` (character): Patient ID. Each patient has 2 libraries which differ in their `condition`
* `condition` (character): Experimental condition. Media for media control. Mtb of *M. tuberculosis* infected.
* `age_dys` (numeric): Age in ages
* `sex` (character): Biological sex, M or F
* `ptID_old` (character): Old patient ID with leading zeroes
* `RNAseq` (logical): If the library has pass-filter RNA-seq data
* `methylation` (logical): If the library has pass-filter methylation data
* `total_seq` (numeric): Total number of pass-filter sequences

## Basic plots
### Boxplot

We will start with a boxplot of the total sequences per library. First, we call a ggplot object. Since we've given it no data or aesthetics, this is a blank plot.

```{r}
ggplot()
```

Then, we give the plot data and specify aesthetics for which parts of the data we want to plot. Now, we see that ggplot knows what the scale of the y data will be. Throughout this section, we will highlight lines of code that change with the comment `#<--- see change`.

```{r}
ggplot(dat$targets) +
  aes(y = total_seq) #<--- see change
```

Finally, we add a geom to denote how we want the data displayed, in this case as a boxplot.

```{r}
ggplot(dat$targets) +
  aes(y = total_seq) +
  geom_boxplot() #<--- see change
```

Since we did not specify an x-variable, all the data are plotted in one boxplot. We add an aesthetic to specify x and split up the data.

```{r}
ggplot(dat$targets) +
  aes(y = total_seq, x = condition) + #<--- see change
  geom_boxplot()
```

### Data ordering with factors

Our x-variable happens to be in the correct alphabetical order with the `Media` controls before `Mtb` infection. However, alphabetical does not always give you want you want. You can force a variable's order by converting it to a factor. Let's switch the `condition` order.

```{r}
dat$targets %>% 
  mutate(condition_fct = factor(condition, 
                            levels = c("Mtb", "Media"))) %>% #<--- see change
  
  ggplot() +
  aes(y = total_seq, x = condition_fct) + 
  geom_boxplot()
```

Notice how we now pipe `%>%` the data into ggplot so that we can modify it in `dplyr` first.
 
### Dotplot

These same data could be displayed as a dotplot instead. We return to the original `condition` order since it was correct.

```{r}
ggplot(dat$targets) +
  aes(y = total_seq, x = condition) +
  geom_point() #<--- see change
```

And if we want to make sure we see every point, a jitter works nicely.

```{r}
ggplot(dat$targets) +
  aes(y = total_seq, x = condition) +
  geom_jitter() #<--- see change
```

### Multiple geoms

The building block nature of ggplot allows us to add multiple geoms on top of each other. So we can combine the two previous plots like so

```{r}
ggplot(dat$targets) +
  aes(y = total_seq, x = condition) +
  geom_boxplot() + #<--- see change
  geom_jitter() #<--- see change
```

But you may notice that there are some duplicate points. Looking back at the first boxplots, there are points for data outside the box. These points remain in the above plot plus appear jittered from the second geom. We can fix this by telling the boxplot to not plot the outliers.

```{r}
ggplot(dat$targets) +
  aes(y = total_seq, x = condition) +
  geom_boxplot(outlier.shape = NA) + #<--- see change
  geom_jitter()
```

#### Layer order matters

Because ggplot works in layers, the order matters. If, for example, we put the boxplot after the jitter, the boxes cover up points! We use `alpha` for transparency so you can see these points. Thus, be careful when considering how you setup your plot!

```{r}
ggplot(dat$targets) +
  aes(y = total_seq, x = condition) +
  geom_jitter() + #<--- see change
  geom_boxplot(outlier.shape = NA, alpha = 0.8)  #<--- see change
```

#### Why are my dots moving?

With jitter, you are adding random error to be able to see all the points. Because this is truly random, every time you run jitter, you will get a slightly different plot. To combat this, set a seed before making the plot.

```{r}
set.seed(468) #<--- see change
ggplot(dat$targets) +
  aes(y = total_seq, x = condition) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter()

set.seed(468) #<--- see change
ggplot(dat$targets) +
  aes(y = total_seq, x = condition) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter()
```

## RNA-seq plots
### PCA

One dotplot of interest in RNA-seq analysis is principle component analysis (PCA). In this plot, each RNA-seq library is a single point. The axes are unitless but in general, points that are closer together represent libraries with more similar gene expression. You can see large scale changes in gene expression in PCA. However, even when you don't see clear PCA groupings, there can be significant differentially expressed genes. So don't despair if your data don't cluster in this plot!

Here, we calculate PCA from the normalized counts. Note that we transpose `t()` the counts data to get one PCA point per library, instead of one per gene.

```{r}
dat.pca <- prcomp(t(dat$E), scale.=TRUE, center=TRUE)

dat.pca$x[1:3,1:3]
```

Then we use join to add metadata to the PCA data, so we can annotate the PCA plot with variables of interest.

```{r}
dat.pca.meta <- as.data.frame(dat.pca$x) %>% 
  rownames_to_column("libID") %>% 
  inner_join(dat$targets)
```

A nice trick with joins is to not specify what to join `by`. In this case, the function finds all columns shared between the data frames and uses them to match rows.

With `geom_point`, we have a PCA!

```{r}
ggplot(dat.pca.meta) +
  aes(x = PC1, y = PC2) +
  geom_point()
```

#### Color and shape

But it's not very informative, so let's annotate with our condition groups. In general, you want to use annotations in the following order: space, color, shape, size. These are ordered by how easily the human eye can discern differences.

We already have space as libraries are separated along x and y. Now we add color.

```{r}
ggplot(dat.pca.meta) +
  aes(x = PC1, y = PC2, color = condition) + #<--- see change
  geom_point()
```

I like to keep all my aesthetics in one layer, but you could keep adding `aes()` functions to the plot if you prefer. This gives the same plot.

```{r eval=FALSE}
ggplot(dat.pca.meta) +
  aes(x = PC1, y = PC2) + #<--- see change
  geom_point() +
  aes(color = condition) #<--- see change
```

We could also use shape for the same data. Which plot to you think is easier to interpret?

```{r}
ggplot(dat.pca.meta) +
  aes(x = PC1, y = PC2, shape = condition) + #<--- see change
  geom_point()
```

We're not showing size or transparency because these aesthetics are difficult to see and should be avoided unless you *really* need them or the differences between data points are large enough to see in these aesthetics.

#### Labels

We can also improve our axes and legend labels. First, we see what proportion of variation is explained by each PC.

```{r}
summary(dat.pca)$importance
```

And use that for labels as well as change the legend.

```{r}
ggplot(dat.pca.meta) +
  aes(x = PC1, y = PC2, color = condition) +
  geom_point() +
  labs(x = "PC1 (24%)", y = "PC2 (11%)", #<--- see change
       color = "Infection")
```

#### Coordinates

Since PCA units are relative, it is nice to have the scales of the x and y be the same. That is, 1 inch of space corresponding to the same values for both. This is accomplished with `coord_fixed`. Note that in this PCA, this function does not change the plot much but it can can a big effect in other PCA! There are also several other coordinate systems you can see with `coord_`.

```{r}
ggplot(dat.pca.meta) +
  aes(x = PC1, y = PC2, color = condition) +
  geom_point() +
  labs(x = "PC1 (24%)", y = "PC2 (11%)",
       color = "Infection") +
  coord_fixed() #<--- see change
```

#### Themes

`ggplot` has a number of themes that change the aesthetics of the plot. I like `theme_classic` for PCA. Type `theme_` into the console to see the available themes suggested with auto-complete. You can also checkout more themes in the package `ggthemes`.

```{r}
ggplot(dat.pca.meta) +
  aes(x = PC1, y = PC2, color = condition) +
  geom_point() +
  labs(x = "PC1 (24%)", y = "PC2 (11%)",
       color = "Infection") +
  coord_fixed() +
  theme_classic() #<--- see change
```

### volcano plot

Let's move on to another application of `geom_point`. A volcano plot shows differentially expressed gene significance and fold change.

#### Linear modeling

We'll need to perform linear modeling to determine these results. Below is the workflow for comparing Mtb-infected vs media samples, corrected for sex and blocked by patient. We use our R package [`kimma`](https://github.com/BIGslu/kimma). You **DO NOT** need to run these steps. They are here in case you want to run a similar analysis in the future.

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("BIGslu/kimma")
```

```{r eval=FALSE}
library(kimma)
```

```{r eval=FALSE}
fit <- kmFit(dat, model = "~ condition + sex + (1|ptID)",
             run.lme = TRUE, use.weights = TRUE)
```

```
lme/lmerel model: expression~condition+sex+(1|ptID)
Input: 20 libraries from 10 unique patients
Model: 20 libraries
Complete: 13419 genes
Failed: 0 genes
```

```{r eval=FALSE}
save(fit, file="data/kimma_result.RData")
```

Instead, load the results from the data downloaded for this workshop. Note that this is an S3 object similar to the `dat` object. In it, we have the linear mixed effects (lme) model results in `fit$lme`, which includes our log2 fold change `estimate` and significance `FDR` for each gene and each variable.

```{r}
load("data/kimma_result.RData")

fit$lme
```

#### Color

Let's start with a basic dotplot like before. We first filter the fit data to just the variable `condition` so that we have 1 point per gene.

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR)) +
  geom_point()
```

Now let's color the significant points. There is no "significant" variable in the data frame, so we first make one with mutate. 

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 ~ "FDR < 0.01", #<--- see change
                                  TRUE ~ "NS")) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR), color = significance) + #<--- see change
  geom_point()
```

Or go further and color up vs down regulated genes differently.

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down", #<--- see change
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR), color = significance) +
  geom_point()
```

And we can specify the colors we want with another layer.

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR), color = significance) +
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red")) #<--- see change
```

#### Lines

Since we're defining significance at a cutoff, we can plot that cutoff with a vertical line with `geom_vline`. Other similar geoms are `geom_hline` for a horizontal line and `geom_abline` for a line with slope `a` and intercept `b`.

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR), color = significance) +
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) #<--- see change
```

#### Scales

You might also wish to center the 0 estimate or otherwise alter the plots limits.

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR), color = significance) +
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) +
  lims(x = c(-9, 9)) #<--- see change
```

If you accidentally (or purposefully) remove data point with your limits, R will warn you!

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR), color = significance) +
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) +
  lims(x = c(-3, 3)) #<--- see change
```

#### Label points

We're going to use a ggplot extension package to label our most significant genes. Let's load that package now.

```{r}
library(ggrepel)
```

First, we determine our top 2 up and down regulated genes. There are some new `dplyr` functions in here! See the comments for information.

```{r}
top_deg <- fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(direction = case_when(estimate < 0 ~ "down",
                               estimate > 0 ~ "up")) %>% 
  group_by(direction) %>% 
  slice_min(FDR, n=2) %>% # Keeps n number of minimum FDR values
  pull(gene) # Extract the variable and turns the data into a vector

top_deg
```

Then we add those labels to our fit data and use that variable in `geom_text_repel`. In this case, the data removal warning is expected as we're not labeling the majority of points.

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  mutate(label = case_when(gene %in% top_deg ~ gene)) %>%  #<--- see change
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR), color = significance) +
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) +
  lims(x = c(-9, 9)) +
  geom_text_repel(aes(label = label)) #<--- see change
```

The text legend gets overlayed with the point legend, which looks odd. Let's remove one of them.

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  mutate(label = case_when(gene %in% top_deg ~ gene)) %>%
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR), color = significance) +
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) +
  lims(x = c(-9, 9)) +
  geom_text_repel(aes(label = label), show.legend = FALSE) #<--- see change
```

And a small edit if you don't want the text to be colored. When you put an aesthetic in a specific geom, it only applies that change to that one geom.

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  mutate(label = case_when(gene %in% top_deg ~ gene)) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR)) + #<--- see change
  geom_point(aes(color = significance)) + # <--- see change
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) +
  lims(x = c(-9, 9)) +
  geom_text_repel(aes(label = label), show.legend = FALSE)
```

#### Theme

To round it out, let's look at `theme_bw`, my favorite one for volcano plots.

```{r}
fit$lme %>% 
  filter(variable == "condition") %>% 
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  mutate(label = case_when(gene %in% top_deg ~ gene)) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR)) +
  geom_point(aes(color = significance)) + 
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) +
  lims(x = c(-9, 9)) +
  geom_text_repel(aes(label = label)) +
  theme_bw() #<--- see change
```

There you have it! Did you think you'd reach a 11 function plot so soon? 

#### Facets

With big data, we often want to plot more than one thing at once. Facets allow you to make similar plots across multiple subsets of data with only 1 additional line of code. Here, we filter for 2 `variable` and then facet those into 2 plots.

```{r}
fit$lme %>% 
  filter(variable %in% c("condition", "sex")) %>% #<--- see change
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  mutate(label = case_when(gene %in% top_deg ~ gene)) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR)) +
  geom_point(aes(color = significance)) + 
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) +
  #<--- see change No lims used
  geom_text_repel(aes(label = label)) +
  theme_bw() +
  facet_wrap(~ variable) #<--- see change
```

Since the a and y scales differ for the two variable, we allow them to differ with `scales`. This is my we removed the `lims` function earlier. `scales` can be set to "free_x", "free_y", or "free" for both x and y.

```{r}
fit$lme %>% 
  filter(variable %in% c("condition", "sex")) %>%
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  mutate(label = case_when(gene %in% top_deg ~ gene)) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR)) +
  geom_point(aes(color = significance)) + 
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) +
  geom_text_repel(aes(label = label)) +
  theme_bw() +
  facet_wrap(~ variable, scales = "free") #<--- see change
```

### Save plots

Once you have your plot as you desire, save it to your computer for use in presentations, publications, Twitter, and more! `ggsave` allows you to save in a bunch of formats (see the help page with `?ggsave`). 

Since we've only been printing our plots to the plot pane in RStudio, we need to first save a plot to the environment. We'll use the single `condtion` variable volcano plot.

```{r}
p1 <- fit$lme %>% #<--- see change
  filter(variable == "condition") %>%
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  mutate(label = case_when(gene %in% top_deg ~ gene)) %>% 
  
  ggplot() +
  aes(x = estimate, y = -log10(FDR)) +
  geom_point(aes(color = significance)) + 
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_hline(yintercept = -log10(0.01)) +
  geom_text_repel(aes(label = label)) +
  theme_bw() 

p1
```

Now we can save the object. My most commonly used plot formats are `.pdf` for publication (vectorized PDF that can be edited in Illustrator/Inkscape) and `.png` for all other uses (not vectorized, smaller files). You specify the file type in the filename and give dimensions in inches (default). Without dimensions, R used whatever the ratio of you plot pane currently is (not reproducible!).

```{r}
ggsave(filename = "images/volcano_plot.png", plot = p1, width = 3, height = 4)
ggsave(filename = "images/volcano_plot.pdf", plot = p1, width = 3, height = 4)
```

## Commented plot scripts

Here are each of the final plots fulled comments for your reference.

```{r eval = FALSE}
ggplot(dat.pca.meta) +                   # Make ggplot from data in dat.pca.meta
  aes(x = PC1, y = PC2,                  # Plot PCs 1 and 2 along x and y
      color = condition) +               # Color by condition
  geom_point() +                         # Plot points for each value
  labs(x = "PC1 (24%)", y = "PC2 (11%)", # Relabel axes
       color = "Infection") +            # Relabel legend
  coord_fixed() +                        # Force xy scales to be equal
  theme_classic()                        # Set theme
```

```{r eval = FALSE}
fit$lme %>% 
  # Keep results for only the condition variable
  filter(variable == "condition") %>% 
  # Create variable for significant up and down genes
  mutate(significance = case_when(FDR < 0.01 & estimate < 0 ~ "down",
                                  FDR < 0.01 & estimate > 0 ~ "up",
                                  TRUE ~ "NS")) %>% 
  # Create variable for gene name labels
  mutate(label = case_when(gene %in% top_deg ~ gene)) %>% 
  
  ggplot() +                                              # initiate ggplot plot
  aes(x = estimate, y = -log10(FDR)) +                    # plot estimate and FDR
  geom_point(aes(color = significance)) +                 # plot points, color by significance
  scale_color_manual(values = c("blue", "grey", "red")) + # recolor points
  geom_hline(yintercept = -log10(0.01)) +                 # add FDR cuttof line
  lims(x = c(-9, 9)) +                                    # set axis limits
  geom_text_repel(aes(label = label)) +                   # label points
  theme_bw()                                              # set theme
```

```{r eval = FALSE}
  ... + 
  facet_wrap(~ variable) # facet by variable
```

## Exercises: ggplot

1. Using the combined expression and metadata you made in the last session, plot the expression of one gene in `Media` vs `Mtb`. The plot type is up to you! As a reminder, here is how we made the combined data.

```{r}
full_data <- as.data.frame(dat$E) %>% 
  rownames_to_column("gene") %>% 
  pivot_longer(-gene, names_to = "libID", values_to = "log2CPM") %>% 
  inner_join(dat$targets)

full_data
```

2. Color the PCA by the numeric variable `total_seq`. Explore the help pages for `scale_color_continuous`, `scale_color_gradient`, or `scale_color_gradient2` and recolor your PCA to more easily see differences. As a reminder, here is the base PCA plot.

```{r}
ggplot(dat.pca.meta) +
  aes(x = PC1, y = PC2) +
  geom_point() +
  labs(x = "PC1 (24%)", y = "PC2 (11%)",
       color = "Infection") +
  theme_classic()
```

3. Using `geom_text_repel`, label the points in the PCA by their `libID`

# *Navigation*

* [Workshop index][index]
* Previous lesson: [Data visualization in ggplot][lesson3]
* [Lesson source code][lesson3rmd]

***

[index]: https://bigslu.github.io/workshops/2022.08.15_R.tidyverse.workshop/index.html
[lesson0]: https://bigslu.github.io/workshops/2022.08.15_R.tidyverse.workshop/0_setup.html
[lesson1]: https://bigslu.github.io/workshops/2022.08.15_R.tidyverse.workshop/1_introR.html
[lesson2]: https://bigslu.github.io/workshops/2022.08.15_R.tidyverse.workshop/2_tidyverse.html
[lesson3]: https://bigslu.github.io/workshops/2022.08.15_R.tidyverse.workshop/3_ggplot.html
[lesson1rmd]: https://github.com/BIGslu/workshops/blob/main/2022.08.15_R.tidyverse.workshop/1_introR.Rmd
[lesson2rmd]: https://github.com/BIGslu/workshops/blob/main/2022.08.15_R.tidyverse.workshop/2_tidyverse.Rmd
[lesson3rmd]: https://github.com/BIGslu/workshops/blob/main/2022.08.15_R.tidyverse.workshop/3_ggplot.Rmd

[workshop]: https://bigslu.github.io/workshops/2022.08.15_R.tidyverse.workshop/R_tidyverse_RNAseq_edition.html
