---
title: "Customizing ggplot"
author: "Kim Dill-McFarland, kadm@uw.edu"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
urlcolor: blue
---
```{r include=FALSE}
knitr::opts_chunk$set(fig.width = 8.5, fig.height = 3.5)
```

# Overview
In this workshop, we explore some options to customize ggplot. This document contains the basic plot to start and then addtional plots with modifications in response to attendee questions.

# Prior to the workshop

Please install [R](http://www.r-project.org), [RStudio](https://www.rstudio.com/products/rstudio/download/), and the following packages.

```{r message=FALSE, warning=FALSE}
#Data manipulation
#install.packages("tidyverse")
library(tidyverse)
#Example data
#install.packages("devtools")
#devtools::install_github("BIGslu/kimma")
library(kimma)
```

# Load data

Briefly, these data are from RNA-sequencing of human dendritic cells cultured with and without virus. Samples are from 3 donors and a random subset of 1000 genes were selected. Expression data are in an limma EList object (named `example.voom`) containing expression (`E`), sample/patient metadata (`targets`), and gene metadata (`genes`). Expression is expressed as TMM-normalized log2 counts per million (CPM).

We combine the expression, sample, and gene metadata in a single table for use in plotting.

```{r message=FALSE}
dat <- as.data.frame(example.voom$E) %>% 
  rownames_to_column("geneName") %>% 
  pivot_longer(-geneName, names_to = "libID") %>% 
  inner_join(example.voom$targets, by = "libID") %>% 
  inner_join(example.voom$genes, by = "geneName")

dat
```

Additionally, we load table with GSEA results for these data. See the `example_gsea.R` script for the code to run GSEA.

```{r}
gsea <- read_csv("data/example_gsea.csv", show_col_types = FALSE)
class(gsea)
gsea
```

# Plots
## Boxplot

```{r}
dat %>% 
  filter(hgnc_symbol == "IFIT3") %>% 
  
  ggplot(aes(x = virus, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter()
```

Only show some points

```{r}
dat %>% 
  filter(hgnc_symbol == "IFIT3") %>% 
  
  ggplot(aes(x = virus, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(data=filter(dat, 
                          hgnc_symbol == "IFIT3" & donorID=="donor1")) +
  #theme_classic() +
  theme_bw()
```

See `ggthemes` for more

Save to file

```{r}
plot1 <- dat %>% 
  filter(hgnc_symbol == "IFIT3") %>% 
  
  ggplot(aes(x = virus, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter()

ggsave(filename = "plot1.png", plot1,
       width=3, height=3)
```

Width of boxplot relative to sample size.

```{r}
dat %>% 
  filter(hgnc_symbol == "IFIT3") %>% 
  arrange(virus) %>% 
  slice_head(n=9) %>% 
  
  ggplot(aes(x = virus, y = value)) +
  geom_boxplot(varwidth = TRUE,
               outlier.shape = NA) +
  geom_jitter()
```

Violin version with scale to sample size

```{r}
dat %>% 
  filter(hgnc_symbol == "IFIT3") %>% 
  arrange(virus) %>% 
  slice_head(n=9) %>% 
  
  ggplot(aes(x = virus, y = value)) +
  geom_violin(scale = "count")
```

## Barplot

```{r}
dat %>% 
  filter(hgnc_symbol == "IFIT3") %>% 
  
  ggplot(aes(x = donorID, y = value)) +
  geom_bar(aes(fill = virus), stat = "identity",
           position = 'dodge')
```

Stacked bars

```{r}
dat %>% 
  filter(hgnc_symbol == "IFIT3") %>% 
  
  ggplot(aes(x = donorID, y = value)) +
  geom_bar(aes(fill = virus), stat = "identity")
```

```{r}
dat %>% 
  filter(hgnc_symbol == "IFIT3") %>% 
  
  ggplot(aes(x = donorID, y = value)) +
  geom_bar(aes(fill = virus), stat = "identity",
           position = 'dodge') +
  scale_fill_manual(values = c('HRV'='black',
                               'none'='grey60'))
```

Color ideas at <https://colorbrewer2.org/>

Plot mean with error

```{r}
dat %>% 
  filter(hgnc_symbol == "IFIT3") %>% 
  
  ggplot(aes(x = virus, y = value)) +
  geom_bar(aes(fill = virus), 
           stat = "summary", fun = "mean") +
  stat_summary(fun.data="mean_sdl")
```

## "Lollipop" plot

```{r}
gsea %>% 
  filter(padj < 0.3) %>% 
  mutate(Significance = ifelse(padj < 0.2, "FDR < 0.2", "NS")) %>% 
  
  ggplot() +
  geom_segment(aes(x=pathway, xend=pathway, 
                   y=0, yend=NES)) +
  geom_point(aes(x=pathway, y=NES,
                 color = Significance)) +
  geom_hline(yintercept = 0) +
  coord_flip()
```

Sort

```{r}
gsea %>% 
  filter(padj < 0.3) %>% 
  mutate(Significance = ifelse(padj < 0.2, "FDR < 0.2", "NS")) %>% 
  
  ggplot() +
  geom_segment(aes(y=pathway, yend=pathway, 
                   x=0, xend=NES)) +
  geom_point(aes(y=pathway, x=NES,
                 color = Significance)) +
  geom_vline(xintercept = 0)
```

Reorder by NES

```{r}
gsea %>% 
  filter(padj < 0.3) %>% 
  mutate(Significance = ifelse(padj < 0.2, "FDR < 0.2", "NS")) %>% 
  
  ggplot() +
  geom_segment(aes(y=reorder(pathway,NES), yend=reorder(pathway,NES), 
                   x=0, xend=NES)) +
  geom_point(aes(y=reorder(pathway,NES), x=NES,
                 color = Significance)) +
  geom_vline(xintercept = 0)
```

```{r}
gsea %>% 
  filter(padj < 0.3) %>% 
  mutate(Significance = ifelse(padj < 0.2, "FDR < 0.2", "NS")) %>% 
  
  ggplot() +
  geom_segment(aes(y=reorder(pathway,NES),
                   yend=reorder(pathway,NES), 
                   x=0, xend=NES), size=1) +
  geom_point(aes(y=reorder(pathway,NES), x=NES,
                 color = Significance), size=2) +
  geom_vline(xintercept = 0, lty="dashed") +
  theme_bw() +
  theme(legend.position = c(0.1,0.9),
        legend.direction = "vertical",
        plot.title = element_text(hjust = 0.5),
        legend.justification = "left") +
  labs(x="Normalized enrichment score",
       y="", title="Hallmark pathways") +
  lims(x=c(-2.1,2.1))
```

# R session

```{r}
sessionInfo()
```

***